---
output: html_document
---

## Data Preprocessing 

```{r setup, include=FALSE}
library(tidyverse)
library(foreign)
library(haven)
```

### Data importing from NHANES: https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Examination&

```{r data_import}
# Import data from NHANES for 2003-2004
bmx_d1 = read_xpt("data/BMX_C.xpt") |>
  janitor::clean_names() |>
  drop_na(bmxbmi) |>
  select(seqn, bmxbmi)

demo_d1 = read_xpt("data/DEMO_C.xpt") |>
  janitor::clean_names() |>
  filter(ridageyr >= 20 & ridageyr < 85 & dmdmartl != 77 & dmdmartl != 99 & ridexprg %in% c(2,3,NA) & dmdeduc2 != 7 & dmdeduc2 != 9) |>
  select(seqn, riagendr, ridageyr, ridreth1, dmdeduc2, dmdmartl, indfmpir) |>
  drop_na()

paxraw_d1 = read_xpt("data/paxraw_c.xpt") |>
  janitor::clean_names() |>
  filter(paxstat == 1 & paxcal == 1) |>  
  drop_na(paxinten) |>
  select(seqn, paxn, paxinten)


# Import data from NHANES for 2005-2006 
bmx_d2 = read_xpt("data/BMX_D.xpt") |>
  janitor::clean_names() |>
  drop_na(bmxbmi) |>
  select(seqn, bmxbmi)

demo_d2 = read_xpt("data/DEMO_D.xpt") |>
  janitor::clean_names() |>
  filter(ridageyr >= 20 & ridageyr < 85 & dmdmartl != 77 & dmdmartl != 99 & ridexprg %in% c(2,3,NA) & dmdeduc2 != 7 & dmdeduc2 != 9) |>
  select(seqn, riagendr, ridageyr, ridreth1, dmdeduc2, dmdmartl, indfmpir) |>
  drop_na()

paxraw_d2 = read_xpt("data/paxraw_d.xpt", col_select = c("SEQN", "PAXN", "PAXINTEN", "PAXSTAT", "PAXCAL")) |> # only select required columns due to memory limit issue
  janitor::clean_names() |>
  filter(paxstat == 1 & paxcal == 1) |>  
  drop_na(paxinten) |>
  select(seqn, paxn, paxinten)
```


### cleaning into 5 minute interval data 

```{r cleaning as EntroLLM}
# Remove missing values and average the physical activity data over 5-minute intervals for each subject
missing1 = paxraw_d1 |>
  group_by(seqn) |> 
  summarize(n = n()) |>
  filter(n != 24*60*7)

data1 = paxraw_d1 |>
  mutate(min5 = floor((paxn - 1) / 5) + 1) |> # min5 interval added ranging from 1 to 2016
  group_by(seqn, min5) |> 
  summarise(intensity = mean(paxinten)) |> 
  anti_join(missing1, by = "seqn") |>
  inner_join(bmx_d1, by = "seqn") |>
  inner_join(demo_d1, by = "seqn") # 3486 subjects 

data1 |>
  group_by(seqn) |>
  summarize()


missing2 = paxraw_d2 |>
  group_by(seqn) |> 
  summarize(n = n()) |>
  filter(n != 24*60*7)

data2 = paxraw_d2 |>
  mutate(min5 = floor((paxn - 1) / 5) + 1) |> # min5 interval added ranging from 1 to 2016
  group_by(seqn, min5) |> 
  summarise(intensity = mean(paxinten)) |> 
  anti_join(missing2, by = "seqn") |>
  inner_join(bmx_d2, by="seqn") |>
  inner_join(demo_d2, by="seqn") # 3457 subjects

data2 |>
  group_by(seqn) |>
  summarize()

data_total = rbind(data1,data2) |>
  rename(time = min5, 
         gender = riagendr, 
         age = ridageyr, 
         race = ridreth1, 
         education = dmdeduc2, 
         married = dmdmartl, 
         pir = indfmpir) |>
  dplyr::select(seqn, bmxbmi, time, gender, age, race, education, married, pir, intensity) # 6943 subjects 

data_total |>
  group_by(seqn) |>
  summarize()
```

Variables: 
1) BMI = 0: < 25, 1: >= 25 
2) gender = 1: male, 2: female 
3) race = 1: Mexican America, 2: Other Hispanic, 3: Non-Hispanic White, 4: Non-Hispanic Black, 5: Other
4) education = 1: less than 9th grade, 2: 9-11th grade (includes 12th grade and no diploma), 3: high school grad/GED or equivalent, 4: some college or associates (AA) degree, 5: college graduate or higher 
5) marital_status = 1: married, 2: widowed, 3: divorced, 4: separated, 5: never married, 6: living with partner 
6) PA (physical activity) = 0: intensity < 100 counts/min (sedentary), 1: 100 <= intensity < 760 counts/min (light), 2: 760 <= intensity < 2200 counts/min (lifestyle), 3: 2200 <= intensity < 6000 counts/min (moderate), 4: intensity >= 6000 counts/min (vigorous)

```{r}
data_wide = data_total |> 
  mutate(
    bmi = ifelse(bmxbmi >= 25, 1, 0),
    pa = ifelse(intensity < 100, 0,
                ifelse(intensity < 760, 1,
                       ifelse(intensity < 2200, 2,
                              ifelse(intensity < 6000, 3, 4))))) |>
  select(-bmxbmi, -intensity) |>
  pivot_wider(
    names_from = time, 
    values_from = pa, 
    names_prefix = "time")

save(data_wide, file = "data/data_wide.RData")
write.csv(data_wide, file = "data/data_wide.csv")       
```



### 20 minute interval data 

```{r}
data120 = paxraw_d1 |>
  mutate(min20 = floor((paxn - 1) / 20) + 1) |> # min20 interval added ranging from 1 to 2016
  group_by(seqn, min20) |> 
  summarise(intensity = mean(paxinten)) |> 
  anti_join(missing1, by = "seqn") |>
  inner_join(bmx_d1, by = "seqn") |>
  inner_join(demo_d1, by = "seqn") # 3486 subjects 

data220 = paxraw_d2 |>
  mutate(min20 = floor((paxn - 1) / 20) + 1) |> # min20 interval added ranging from 1 to 2016
  group_by(seqn, min20) |> 
  summarise(intensity = mean(paxinten)) |> 
  anti_join(missing2, by = "seqn") |>
  inner_join(bmx_d2, by="seqn") |>
  inner_join(demo_d2, by="seqn") # 3457 subjects

data_total_20 = rbind(data120,data220) |>
  rename(time = min20, 
         gender = riagendr, 
         age = ridageyr, 
         race = ridreth1, 
         education = dmdeduc2, 
         married = dmdmartl, 
         pir = indfmpir) |>
  dplyr::select(seqn, bmxbmi, time, gender, age, race, education, married, pir, intensity) # 6943 subjects 


data_wide_20 = data_total_20 |> 
  mutate(
    bmi = ifelse(bmxbmi >= 25, 1, 0),
    pa = ifelse(intensity < 100, 0,
                ifelse(intensity < 760, 1,
                       ifelse(intensity < 2200, 2,
                              ifelse(intensity < 6000, 3, 4))))) |>
  select(-bmxbmi, -intensity) |>
  pivot_wider(
    names_from = time, 
    values_from = pa, 
    names_prefix = "time")

save(data_wide_20, file = "data/data_wide_20.RData")
write.csv(data_wide_20, file = "data/data_wide_20.csv")    
```


## Visualization of data 

```{r}
library(ggplot2)

# Choose a few patient IDs randomly
sample_id = sample(unique(data_total$seqn), 20)

# Filter data for those patients
sample_data = data_total %>% filter(seqn %in% sample_id)

# Plot number of diagnoses over time for each patient
ggplot(sample_data, aes(x = time, y = intensity, group = seqn)) +
  geom_line() +
  facet_wrap(~seqn, scales = "free_y") +
  labs(title = "Weekly Diagnosis Count per Patient", x = "Week", y = "Diagnosis Count")

```


